Motion Aftereffects in the vestibular system: influence of (spatially congruent) subsequent own-body motion on self-motion perception

### 2017-07-27 manuscript.Rmd
first long version of manuscript

---
title: "Reversed motion direction illusion: influence of (spatially congruent) subsequent own-body motion on self-motion perception"
shorttitle: "Reversed motion direction illusion"
author:
  - name: Roman B. Di Francesco
    affiliation: 1
  - name: Andrew W. Ellis
    affiliation: 1
  - name: Luzia Grabherr  
    affiliation: 1
  - name: Gianluca Macauda
    corresponding: yes    # Define only one corresponding author
    address: Departement of Psychology, University of Bern, Fabrikstrasse 8, 3012 Bern, Switzerland
    email: gianluca.macauda@psy.unibe.ch
  - name: Fred W. Mast
    affiliation: 1
affiliation:
  - id: 1
    institution: Departement of Psychology, University of Bern

abstract: |
  Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet.

Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse molestie consequat, vel illum dolore eu feugiat nulla facilisis at vero eros et accumsan et iusto.

@@

  plot_correct <- exp_all_correct %>%
  ggplot(aes(x = soa, y = meanCorrect, shape = cue, color = cue)) + 
  geom_hline(yintercept = .5, linetype = 'dashed') +
  stat_summary(fun.y = mean, geom = 'point') + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar") + 
  #geom_boxplot() +
  coord_cartesian(ylim = c(0,1), expand = 0) + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
       title = 'Accuracy over all Conditions', 
       color = 'Motion Congruence', 
       shape = 'Motion Congruence') + 
  ylab('Accuracy') + 
  xlab('Inter-Stimulus-Intervall') +
  facet_grid(~experiment) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
        axis.title = element_text(size = 12, face = 'bold'), 
        legend.title = element_text(size = 12, face = 'bold'), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

@@

load('figures_ddm.Rda')
```
# Introduction
The malleable nature of human perception is best observed in sensory illusions. One of such illusions is known as motion aftereffect (MAE): Classically, in the MAE a prolonged exposure to moving visual stimulus leads to the temoporarily illusory perception of motion in the other opposite direction of a subsequent static stimulus. Famously described by Adams as the waterfall illusion [@adams_account_1834], since then the MAE has become an extremely interesting paradigm in vision science. Albeit their simplicity, MAEs are a powerful tool to investigate underlying neural and computational mechanisms of, mainly visual, motion perception [@anstis_motion_1998; @cuturi_optic_2014; @konkle_motion_2009]. There, the dominating explanation models rely on selective adaptation of motion-sensitive visual neurons. In such models, a prolonged exposition to a specific motion leads to a reduced firing rate and responiveness of those neurons. Assuming that distinct populations of neurons code different directions which are constantly compared, a prolonged stimulation with one direction would result in a decreased firing rate and thus in the perception of apparent motion in the opposite direction [@anstis_motion_1998].

However, MAEs are not limited to the visual modality and similar effects have also been observed in the auditory [@bestelmeyer_auditory_2010; @latinus_perceptual_2012; @reinhardt-rutland_spectrally_2001], proprioceptive [@seizova-cajic_proprioceptive_2007] and recently also vestibular modality [@coniglio_human_2014; @crane_foreaft_2012; @crane_roll_2012]. The modality independent nature of the MAE is further highlighted in crossmodal aftereffects. While there are multisensory effects in the visuo-auditory [@berger_auditory_2016; @ehrenstein_cross-modal_1996] and visuo-tactile domain [@konkle_motion_2009], for the present study visuo-vestibular aftereffects are more relevant.

Studies on visuo-vestibular aftereffects make use of a special kind of visual stimulation, i.e. vection. In vection, visual motion stimuli elicits the illusory feeling of self motion [@brandt_motion_1974]. In this context it has been shown, that vection results in vection after effects [@patterson_modeling_2009; @seno_vection_2010]. Additionally, vestibular information seems to modulate the aftereffects induced by vection [@wallach_compensation_1975]. In their seminal study, Cuturi and MacNeilage [@cuturi_optic_2014] observed that that sustained optic flow induced aftereffects in a subsequent vestibular motion detection task and thus demonstrated crossmodal MAE between the visual and vestibular modality that may rely on different neurophysiological mechanisms than the visual MAE. Taken together, these studies point out that aftereffects are not limited to visual motion in general but also specific for self-motion, which are found in the visual system alone in vection paradigms, in the vestibular system alone [@coniglio_human_2014; @crane_foreaft_2012; @crane_roll_2012] and in the combination of both [@cuturi_optic_2014]. Thus, MAEs seem to be a sensory modality independent phenomenon and an explanatory models limited to lower-level visual areas fail to grasp the crossmodal nature of the sensory motion aftereffect.

For the vestibular system, possible MAEs are highly interesting and important as the vestibular organs track the position of the self relative to space and are thus highly involved in the perception of self motion. Anecdotal evidence for such vestibular MAEs are found in the 'Gillingham illusion' [@ercoline_post-roll_2000], in which pilots tend to counter previous roll motions with a roll motion in the same direction. Driven by this phenomenon, Crane and colleagues found such aftereffects for fore-aft translation [@crane_foreaft_2012], roll [@crane_roll_2012] and yaw rotation [@coniglio_human_2014]. This also highlights that vestibular aftereffects can be evoked in the semicircular canals as well as in the otoliths, giving further evidence for the crossmodal nature of motion aftereffects.

Taken together, these findings contribute to the idea that there is a centre for self-motion perception that integrates information from different modalities [@seno_vection_2010]. However, in the vestibular system an important distinction has to be made between the perception of rotation, which is coded in the semicircular canals, and the perception of translation, which is coded in the otoliths, although their signals are combined at an early processing stage [@crane_perception_2016]. Thus, humans percept of selfmotion is multisensory even within the vestibular system. In every-day life those two signals are rarely perceived alone, which makes crossmodal aftereffects within the vestibular system more interesting.

In the present study, we set out to explore this idea even more with data from three experiments (unpublished data) that tried to shift attention with natural vestibular stimulation consisting of a prior yaw rotation or inertial translation to facilitate the discrimination of the direction of a subsequent congruent or incongruent yaw rotation or inertial translation. Thus, in experiment one a inertial translation was always followed by a yaw rotation in the same or opposite direction, and in experiment two and three a yaw rotation was followed by a inertial translation in the same or opposite direction. Participants' task was always to indicate the perceived direction of the second motion as fast as possible. Thus, the three experiments investigate whether there is a crossmodal aftereffect within the vestibular system and whether it is unilateral.

The structure of the three experiments with the discrimination tasks is ideal to study the perceptual decision making process. As far as we know, the decision making process has never been investigated more closely in vestibular motion aftereffects. However, a joint analysis of response and response time would provide more detailed information about the dynamics of the underlying process. A commonly used model for a joint analysis is the Drift Diffusion Model (DDM) [@ratcliff_diffusion_2008]. In the DDM it is assumed that the decision making process relies on the acculumation of noisy sensory evidence over time. A response is given when a threshold for one of the two possibilities is reached. Several parameters depict the decision making process: The drift rate describes the speed of evidence accumulation for one of the two response options. The starting point indicates a point within the boundary from which evidence accumulation starts. Parameters of interest for the current study are the drift rate and starting point. Other important parameters of the DDM include the boundary separation parameter, which serves as the threshold for the needed evidence to make decision and the non deicision time parameter that covers the portion within a response not involved in the deicision making process (e.g. motor execution). Other studies within the field of perceptual decision making [@ellis_vestibular_2017; @wiech_influence_2014] used DDM to disentagle influence of prior information on different parameters of the decision making process. Applied to MAEs, a first motion stimulation (in our case vestibular stimulation) influences the judgment of the subsequent motion discrimantion task by introducing a bias. However, it remains to be seen whether such a bias shifts the starting point to the opposite direction, which would mean that a first motion alters the perceptual decision making process, or whether the first motion affects the drift rate and thus changes the acculumation process, which would be indicative for altered sensory processing [@wiech_influence_2014]. A combination of both biases is possible as well (see also a simplified depicition of the DDM in  ```\@ref(fig:ddms)```).

The goal of this study is twofold: First, we want to replicate the previously found MAE in the vestibular modality. However, here it tested whether a stimulation of the otliths can evoke MAE for the semicircular canals and vice versa. This would further highlight the crossmodal nature of MAEs. Second and based on the MAE, it is investigated whether the introduced bias for the opposite direction is more likely due to changes in sensory processing or a more general cognitive strategy with the help of a joint analysis of response times and responses in a DDM.

```{r ddms, fig.cap = 'Three possibilites how the first motion could influence the discrimination of the second motion. On the left, the perceptual decision making process would be changed.  The first motion would influence just the starting point. A left motion would shift the starting point towards the decision threshold for a response to the right and vice versa. Here, the drift rate would not be influenced by the first motion. In the middle, altered sensory processing is depicted. The starting point would remain unchanged, the aftereffect would solely be caused by an alteration in the acculumuation process, i.e. a variable drift rate. On the right, the combination of both is presented. The aftereffect would be explained by a shifted starting point as well as a variable drift rate.'}

@@

Experiment 1 was originally planed as a modification of the Spatial Cueing paradigm by Posner et al. (1980) to examine potential influences of own-body motion stimuli on the perception of subsequent motion stimuli. Similar to Spatial Cueing in the visual system, a cue stimulus was expected to direct covert attention to one side in space.  The spatial cueing effect is usually found for short stimuli with short ISIs (Ruz and Lupianez 2002). Therefore both movements were short (200 ms) and the ISIs ranged from 50 ms to 600 ms.

Participants were first translated in the y-axis (cue) and were then rotated about an earth-vertical axis (target). The first motion was chosen to be stronger compared to the second one to ensure that participants perceive the cue correctly, since this could potentially influence attention. A trial was considered as congruent when both motions had the same spatial direction, e.g. left translations and counter-clockwise rotation. The first motion was chosen to be 3.6 times greater than the intensity of the threshold established in a preliminary study and similar to published data (Valko et al. 2012) and the second motion was chosen to be 2.6 times greater than the intensity of the thresholds reported by Grabherr et al [@grabherr_vestibular_2008].

@@

Twelve healthy participants participated in the first experiment (four female, mean age 27, range 24-30 years). All participants were right handed according to a Gemran version of the handedness questionnaire by Chapman and Chapman [@chapman_measurement_1987]. One of the participants received a course credit for participation. None of the participants reported a history of relevant neurological, vestibular or attentional disorders. The study was approved by the ethics committee of the University of Bern and all participants gave written informed consent prior to the experiment in accordance with the Declaration of Helsinki.

@@

A six degree of freedom motion base (6DOF2000E, Moog Inc., East Aurora, NY) and in-house software were used to generate the motion stimuli. Cue (antecedent) stimuli consisted of translations with single cycle sinusoidal acceleration and a frequency $f$ of 5 Hz ($a(t) = A\sin(2 \pi ft) = A\sin(\frac{2 \pi t}{T})$, $T = \frac{1}{f}$) along the y-axis (left/right) and the x-axis (forth/back) as used in previous studies (see e.g. Grabherr et al. 2008; Benson et al. 1989; Crane 2012). Acceleration amplitude (A) was set to $0.25 \frac{m}{s^2}$, i.e. a peak velocity of $0.016 \frac{m}{s}$ ($v_{max} = \frac{AT}{\pi}$) and a displacement of 0.0016 m ($\Delta p = \frac{AT^2}{2\pi}$). For the target (subsequent) stimuli the same acceleration profile as for the cue stimuli was used but with yaw rotations about an earth-vertical axis (left/right). The acceleration amplitude was set to $24 \frac{deg}{s^2}$ ($v_{max} = 1.53 \frac{deg}{s}, \Delta p = .153 deg$).

@@

Within each participant Cue Congruency (three levels: congruent, incongruent and neutral) and interstimulus interval (ISI, four levels: 50 ms, 100 ms, 200 ms and 600 ms) were manipulated. In congruent trials y-axis translations were followed by yaw rotations to the corresponding side. In incongruent trials y-axis translations preceeded yaw rotations to the opposite side. Neutral trials consisted of a translation along the x-axis and a subsequent yaw rotation. The combination of the two factors resulted in 32 motion sets. Every motion set was repeated six times per participant. In-house software based on labVIEW (CITE) was used to record participants' response and reaction times.

@@

Participants were seated upright in a car seat with a five-point harness which was mounted on the motion base. Their head was centered and fixated with a helmet. The experiments were conducted in the dark and participants were blindfolded to prevent the processing of visual cues. White noise was delivered to in-ear headphones at approximately 60 dB to mask sounds from the motion base. Participants were instructed to indicate the perceived direction of the second motion as fast as possible by pressing the corresponding button in either their left or right hand. In case of uncertainty, participants had to make their best guess. The first motion was described as a time varying warning cue, which would indicate the beginning of the second motion but not predict its direction. Practice trials were administered until participants understood the task. Once a participant was comfortable with the task, motion sets were presented in three blocks. The length of the breaks between blocks were determined by participants. In total, 192 trials were randomly presented (48 congruent, 48 incongruent, 96 neutral). A trial consisted of a cue stimulus (200 ms), an ISI (50-600 ms), a target stimulus (200 ms), time to respond (max. 2500 ms), return to the origin (1100 ms) and an intertrial interval of about 1000 ms.

@@

For all experiments analyses and figures were computed in R [@R-base] using the ggplot2 [@R-ggplot2], brms [@R-brms] and rstan [@R-rstan] packages. Analyses and data are accessible on the OpenScienceFramework (OSF) (OSFLINK). WRITE SOMEWHERE SPECIFICS OF HMC, SAMPLES AND CHAINS.

To describe data from experiment 1 in more detail, participants' accuracies were analyzed with a multilevel Bayesian logistic regression model with factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms). 
<!---To analyze participants' reaction times a multilevel Bayesian regression model with an ex-Gaussian likelihood function and factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms) was computed. Only correct trials were included as the underlying mechanism leading to a wrong response might differ. Priors were set to the default implemented in brms.-->

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] were calculated in brms [@R-brms]. Neutral conditions were eliminated from the dataset to simplify the statistical models.

In the first model, for the drift rate fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise) were estimated together with the random effects of participants. For the non-decision time and starting point the random effects of particpants were estimated.

@@

The better model was selected based on the leave-one-out cross-validiation (LOO) [@vehtari_practical_2017].

@@

```{r figDescriptiveCorrect, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
plot_correct
```

```{r figDescriptiveRT, echo=FALSE,  cache = FALSE, warning=FALSE}         
rt_times_plot
```

```{r figure_ddm_driftRate, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
figure_ddm_drift
```

```{r fig_ddm_starting, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
figure_ddm_start
```

### Accuracy
The logistic regression (In Fig accuracy plot Experiment 1) showed that the motion discrimination performance is below chance when the prior stimulus was in the same direction, e.g. a left translation follwed by a counter-clockwise yaw rotation. On the other hand, direction discrimination performance is above chance level when a yaw rotation is preceeded by an incongruent or a neutral fore-aft motion. The discrimination performation did not differ meaningfully between the incongruent and neutral motions, as all 95% CIs for the posterior distribution of the different included zero. Overall, the response patterns indicate MAEs, except for ISIs of 600ms.

@@

In the first experiment it has been shown that there are crossmodal MAEs within the vestibular system in terms of direction discrimination performance if a yaw rotation is preceeded by a inertial translation in the same spatial direction. Thus, a manipulation of the otoliths apparently influence a sensory decision making process in the semicircular canals. However, MAEs are only found for short pauses between the two motion stimuli. More importantly, in a hierarchical DDM the MAEs appear to be explained by a combination of a variable drift and a shifted starting point to the opposite direction of the first motion. It remains to be seen whether for the incongruent conditions whether the correct responses are due to correct sensory evidence accumulation of the second motion or whether the evidence accumulation process is also biased by the first motion. The same response pattern for neutral trials seems to support the former explanation.

>>>>

---
title: "Motion Aftereffects in the vestibular system: influence of (spatially congruent) subsequent own-body motion on self-motion perception"
shorttitle: "Reversed motion direction illusion"
author:
  - name: "Roman B. Di Francesco"
    affiliation: "1"
  - name: "Andrew W. Ellis"
    affiliation: "1"
  - name: "Luzia Grabherr"  
    affiliation: "1"
  - name: "Gianluca Macauda"
    affiliation: "1"
    corresponding: yes    # Define only one corresponding author
    address: "Departement of Psychology, University of Bern, Fabrikstrasse 8, 3012 Bern, Switzerland"
    email: "gianluca.macauda@psy.unibe.ch"
  - name: "Fred W. Mast"
    affiliation: "1"
affiliation:
  - id: "1"
    institution: "Departement of Psychology, University of Bern"

author_note: >
  Complete departmental affiliations for each author (note the indentation, if you start a new paragraph).

abstract: >
  Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse molestie consequat, vel illum dolore eu feugiat nulla facilisis at vero eros et accumsan et iusto.

@@

plot_correct <- exp_all_correct %>%
  ggplot(aes(x = soa, y = meanCorrect, shape = cue, color = cue)) + 
  geom_hline(yintercept = .5, linetype = 'dashed') +
  stat_summary(fun.y = mean, geom = 'point') + 
  stat_summary(fun.data = mean_se, geom = "errorbar") + 
  #geom_boxplot() +
  coord_cartesian(ylim = c(0,1), expand = 0) + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
       title = 'Accuracy over all Conditions', 
       color = 'Motion Congruence', 
       shape = 'Motion Congruence') + 
  ylab('Accuracy') + 
  xlab('Inter-Stimulus-Intervall') +
  facet_grid(~experiment) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
        axis.title = element_text(size = 12, face = 'bold'), 
        legend.title = element_text(size = 12, face = 'bold'), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

@@

load('figures_ddm.Rda')
load('figures_ddm_separate.Rda')
```
# Introduction
The malleable nature of human perception is best observed in sensory illusions. One of such illusions is known as motion aftereffect (MAE): Classically, in the MAE a prolonged exposure to a moving visual stimulus leads to the temporary illusory perception of motion to the opposite direction of a subsequent static stimulus. Famously described by Adams as the waterfall illusion [@adams_account_1834], the MAE has evolved into an extremely interesting paradigm for vision science. Albeit their simplicity, MAEs are a powerful tool to investigate underlying neural and computational mechanisms of motion perception [@anstis_motion_1998; @cuturi_optic_2014; @konkle_motion_2009]. In visual sciences, the dominating explanation models rely on a selective adaptation mechanism of motion-sensitive neurons. In such models, a prolonged exposition to a specific motion subsequently leads to a reduced firing rate and responsiveness of those neurons. Assuming that distinct populations of neurons code different directions which are constantly compared, a prolonged stimulation with one direction would result in a decreased firing rate and thus in the perception of apparent motion to the opposite direction of static patterns [@anstis_motion_1998].

However, MAEs are not limited to the visual modality and similar effects have also been observed in the auditory [@bestelmeyer_auditory_2010; @latinus_perceptual_2012; @reinhardt-rutland_spectrally_2001], proprioceptive [@seizova-cajic_proprioceptive_2007] and recently also vestibular sense [@coniglio_human_2014; @crane_foreaft_2012; @crane_roll_2012]. The multisensory nature of the MAE is further highlighted in cross-modal aftereffects. Multisensory effects have been observed in the visuo-auditory [@berger_auditory_2016; @ehrenstein_cross-modal_1996] and visuo-tactile domain [@konkle_motion_2009], but for the present study we will focus visuo-vestibular MAEs.

Studies on visuo-vestibular MAEs make use of a special kind of visual stimulation, i.e. vection. In vection, visual motion stimuli elicit the illusory feeling of self-motion [@brandt_motion_1974]. In this context it has been shown, that vection results in MAEs regarding the self [@patterson_modeling_2009; @seno_vection_2010]. Additionally, vestibular information seems to modulate the aftereffects induced by vection [@wallach_compensation_1975]. In their seminal study, @cuturi_optic_2014 observed that sustained optic flow induced aftereffects in a subsequent vestibular motion detection task and thus demonstrated a cross-modal MAE between the visual and vestibular senses that may rely on different neurophysiological mechanisms than the visual MAE. Taken together, these studies point out that aftereffects are not limited to visual motion in general but also specific for self-motion, which are found in vection paradigms, in the vestibular system alone [@coniglio_human_2014; @crane_foreaft_2012; @crane_roll_2012] and in the combination of the visual and vestibular senses [@cuturi_optic_2014]. Thus, MAEs seem to be a sensory modality independent phenomenon and an explanatory models limited to lower-level visual areas fail to grasp the cross-modal nature of the sensory motion aftereffect.

For the vestibular system, possible MAEs are highly interesting and important as the vestibular organs track the position of the self relative to space and are thus highly involved in the perception of self motion. Anecdotal evidence for such vestibular MAEs is reported in the 'Gillingham illusion' [@ercoline_post-roll_2000], a phenomenon in which pilots tend to counter previous roll motions with a roll motion in the same direction. Driven by this illusion, @crane_foreaft_2012 and colleagues found such aftereffects for fore-aft translations, roll [@crane_roll_2012] and yaw rotations [@coniglio_human_2014]. This also highlights that vestibular aftereffects include both the semicircular canals as well as in the otoliths. Those findings provide further evidence for the cross-modal nature of motion aftereffects.

Taken together, these findings also contribute to the idea that there is a center for self-motion perception that integrates information from different modalities [@seno_vection_2010]. However, in the vestibular system a distinction has to be made between the perception of rotation, which is coded in the semicircular canals, and the perception of translation, which is coded in the otoliths, although their signals are combined at an early processing stage [@crane_perception_2016; @carriot_integration_2015; @dickman_vestibular_2002]. Thus, humans percept of self-motion is multisensory even within the vestibular system. In every-day life those two signals are rarely perceived separately, which makes the possibilty of cross-modal aftereffects within the vestibular system even more interesting.

In the present study, we set out to explore this idea with data from three experiments (unpublished data) that tried to shift attention with natural vestibular stimulation consisting of a prior yaw rotation or translation to facilitate the discrimination of the direction of a subsequent spatially congruent or incongruent yaw rotation or translation. Thus, in experiment 1 a translation was always followed by a yaw rotation in the same or opposite direction, and in experiment 2 and 3 a yaw rotation was followed by a translation to the same or opposite direction. The participants' task was always to indicate the perceived direction of the second motion as fast as possible. We used those three experiments to investigate whether there is a cross-modal MAE within the vestibular system and whether it is unilateral.

The design of the three experiments with the discrimination tasks is ideal to study the perceptual decision making process. As far as we know, the decision making process has never been investigated in MAEs. However, a joint analysis of response and response time would provide more detailed information about the dynamics of the underlying sensory processing.

A commonly used cognitive model for a joint analysis is the Drift Diffusion Model (DDM) [@ratcliff_diffusion_2008]. In the DDM it is assumed that the decision making process relies on the accumulation of noisy sensory evidence over time. A response is given when a threshold for one of the two possibilities is reached. Several parameters specify the decision making process: The drift rate describes the speed of evidence accumulation for one of the two response options. The starting point indicates a point within the boundary from which evidence accumulation starts. Parameters of interest for the current study are the drift rate and starting point. Other important parameters of the DDM include the boundary separation parameter, which serves as the threshold for the needed evidence to make decision and the non decision time parameter that covers the portion within a response not involved in the decision making process (e.g. motor execution).

Other studies within the field of perceptual decision making [@ellis_vestibular_2017; @wiech_influence_2014] used a DDM to investigate the influence of prior information on different parameters of the decision making process and has recently been proposed as model to study the dynamics of vestibular decision making [@merfeld_dynamics_2016] Applied to MAEs, a first motion stimulation (in our case vestibular stimulation) should influence the judgment of the subsequent motion discrimination task by introducing a bias. However, it remains to be seen whether such a bias shifts the starting point to the opposite direction, which would mean that a first motion alters the perceptual decision making process, or whether the first motion affects the drift rate and thus changes the accumulation process, which would be indicative for altered sensory processing [@wiech_influence_2014]. A combination of both biases is possible as well. Those three possbilities are discussed more extensively in a simplified depicition in Figure \@ref(fig:ddms)).

The goal of this study is twofold: First, we want to replicate the previously found MAE in the vestibular modality. However, here it tested whether a stimulation of the otoliths can evoke MAE for the semicircular canals and vice versa. This would further highlight the cross-modal nature of MAEs. Second and based on the MAE, it is investigated whether the introduced bias for the opposite direction is more likely due to changes in sensory processing or a more general cognitive strategy with the help of a joint analysis of response times and responses in a DDM.

```{r ddms, fig.cap = 'Three possibilites how the first motion could influence the discrimination of the second motion to elicit a MAE. On the left, the perceptual decision making process would be changed.  The first motion would influence just the starting point. A left motion would shift the starting point towards the decision threshold for a response to the right and vice versa. Here, the drift rate would not be influenced by the first motion. In the middle, altered sensory processing is depicted. The starting point would remain unchanged, the aftereffect would solely be caused by an alteration in the acculumuation process, i.e. a variable drift rate. On the right, the combination of both is presented. The aftereffect would be explained by a shifted starting point as well as a variable drift rate.'}

@@

Experiment 1 was originally planed as a modification of the Spatial Cueing paradigm by Posner, Snyder and Davidson [@posner_attention_1980] to examine potential influences of own-body motion stimuli on the perception of subsequent motion stimuli. Similar to Spatial Cueing in the visual system, a cue stimulus was expected to direct covert attention to one side in space.  The spatial cueing effect is usually found for short stimuli with short ISIs [@ruz_review_2002]. Therefore both movements were short (200 ms) and the ISIs ranged from 50 ms to 600 ms.

Participants were first translated in the y-axis (cue) and were then rotated about an earth-vertical axis (target). The first motion was chosen to be stronger compared to the second one to ensure that participants perceive the cue correctly, since this could potentially influence attention. A trial was considered as congruent when both motions had the same spatial direction, e.g. left translations and counter-clockwise rotation. The first motion was chosen to be 3.6 times greater than the intensity of the threshold established in a preliminary study and similar to published data [@valko_vestibular_2012] and the second motion was chosen to be 2.6 times greater than the intensity of the thresholds reported by @grabherr_vestibular_2008.

@@

Twelve healthy participants participated in the first experiment (four female, mean age 27, range 24-30 years). All participants were right handed according to a German version of the handedness questionnaire by @chapman_measurement_1987. One of the participants received a course credit for participation. None of the participants reported a history of relevant neurological, vestibular or attentional disorders. The study was approved by the ethics committee of the University of Bern and all participants gave written informed consent prior to the experiment in accordance with the Declaration of Helsinki.

@@

A six degree of freedom motion base (6DOF2000E, Moog Inc., East Aurora, NY) and in-house software were used to generate the motion stimuli. Cue (antecedent) stimuli consisted of translations with single cycle sinusoidal acceleration and a frequency $f$ of 5 Hz ($a(t) = A\sin(2 \pi ft) = A\sin(\frac{2 \pi t}{T})$, $T = \frac{1}{f}$) along the y-axis (left/right) and the x-axis (forth/back) as used in previous studies [see e.g., @benson_thresholds_1989; @crane_foreaft_2012; @grabherr_vestibular_2008]. Acceleration amplitude (A) was set to $0.25 \frac{m}{s^2}$, i.e. a peak velocity of $0.016 \frac{m}{s}$ ($v_{max} = \frac{AT}{\pi}$) and a displacement of 0.0016 m ($\Delta p = \frac{AT^2}{2\pi}$). For the target (subsequent) stimuli the same acceleration profile as for the cue stimuli was used but with yaw rotations about an earth-vertical axis (left/right). The acceleration amplitude was set to $24 \frac{deg}{s^2}$ ($v_{max} = 1.53 \frac{deg}{s}, \Delta p = .153 deg$).

@@

Within each participant Cue Congruence (three levels: congruent, incongruent and neutral) and inter-stimulus interval (ISI, four levels: 50 ms, 100 ms, 200 ms and 600 ms) were manipulated. In congruent trials y-axis translations were followed by yaw rotations to the corresponding side. In incongruent trials y-axis translations preceded yaw rotations to the opposite side. Neutral trials consisted of a translation along the x-axis and a subsequent yaw rotation. The combination of the two factors resulted in 32 motion sets. Every motion set was repeated six times per participant. In-house software based on labVIEW (National Instruments, Austin, Texas) was used to record participants' response and reaction times.

@@

Participants were seated upright in a car seat with a five-point harness which was mounted on the motion base. Their head was centered and fixated with a helmet. The experiments were conducted in the dark and participants were blindfolded to prevent the processing of visual cues. White noise was delivered to in-ear headphones at approximately 60 dB to mask sounds from the motion base. Participants were instructed to indicate the perceived direction of the second motion as fast as possible by pressing the corresponding button in either their left or right hand. In case of uncertainty, participants had to make their best guess. The first motion was described as a time varying warning cue, which would indicate the beginning of the second motion but not predict its direction. Practice trials were administered until participants understood the task. Once a participant was comfortable with the task, motion sets were presented in three blocks. The length of the breaks between blocks were determined by participants. In total, 192 trials were randomly presented (48 congruent, 48 incongruent, 96 neutral). A trial consisted of a cue stimulus (200 ms), an ISI (50-600 ms), a target stimulus (200 ms), time to respond (max. 2500 ms), return to the origin (1100 ms) and an  interval of about 1000 ms.

@@

For all experiments analyses and figures were computed in R [@R-base] using the ggplot2 [@R-ggplot2], brms [@R-brms] and rstan [@R-rstan] packages. Analyses, models and data are freely accessible on the OpenScienceFramework (OSF) (https://osf.io/46nqw/).

To describe data from experiment 1 in more detail, participants' performances were analyzed with a multilevel Bayesian logistic regression model with factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms). 
<!---To analyze participants' reaction times a multilevel Bayesian regression model with an ex-Gaussian likelihood function and factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms) was computed. Only correct trials were included as the underlying mechanism leading to a wrong response might differ. Priors were set to the default implemented in brms.-->

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] in a Bayesian framework were calculated in brms [@R-brms]. In a hierarchical model group and participant parameters can be estimated at the same time based on the assumption that participants' parameters are drawn from a group distribution [@vandekerckhove_hierarchical_2011; @wiecki_hddm:_2013]. It allows parameter estimates for groups while taking into account variation between participants. Additionally, the Bayesian implementation allows inference of the complete posterior distributions of the parameters of interest. For the first experiment, neutral conditions were left out to simplify the statistical models.

@@

The better model was selected based on the leave-one-out cross-validation (LOO) [@vehtari_practical_2017].

@@

```{r figDescriptiveCorrect, fig.cap='The mean of the proportion of correct answers in the direction discrimination task for all participants in all conditions and experiments. The means over all participants are represented by red dots for congruent conditions, green triangles for incongruent conditions and the blue squares for neutral condtions (only experiment 1). The error bars represent the Standard Error. The dotted line marks the chance level.',  echo=FALSE, cache = FALSE, warning=FALSE}
plot_correct
```

```{r figDescriptiveRT, fig.cap='The mean of the proportion of correct answers in the direction discrimination task for all participants in all conditions and experiments. The means over all participants are represented by red dots for congruent conditions, green triangles for incongruent conditions and the blue squares for neutral condtions (only experiment 1). The error bars represent the Standard Error. The dotted line marks the chance level.',  echo=FALSE, cache = FALSE, warning=FALSE}         
rt_times_plot
```

The logistic regression showed that the motion discrimination performance is below chance when the prior stimulus was in the same direction, e.g. a left translation followed by a counter-clockwise yaw rotation. On the other hand, direction discrimination performance is above chance level when a yaw rotation is preceded by an incongruent or a neutral fore-aft motion (Figure \@ref(fig:figDescriptiveCorrect) - experiment 1). The discrimination performance did not differ meaningfully between the incongruent and neutral motions, as all 95% CIs for the posterior distribution of the different included zero. Overall, the response patterns indicate MAEs, except for ISIs of 600ms.

@@

In the first experiment it has been shown that there are cross-modal MAEs within the vestibular system in terms of direction discrimination performance if a yaw rotation is preceded by a  translation in the same spatial direction. Thus, a manipulation of the otoliths apparently influence a sensory decision making process in the semicircular canals. However, MAEs are only found for short pauses between the two motion stimuli. More importantly, in a hierarchical DDM the MAEs appear to be explained by a combination of a variable drift and a shifted starting point to the opposite direction of the first motion. It remains to be seen whether for the incongruent conditions whether the correct responses are due to correct sensory evidence accumulation of the second motion or whether the evidence accumulation process is also biased by the first motion. The same response pattern for neutral trials seems to support the former explanation.
